---
title: "An Analysis of Personal Movement Using Activity Monitoring Devices"
author: "Janrey Nevado"
date: "5/16/2019"
output: html_document
---

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](http://www.fitbit.com/), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone Up](https://jawbone.com/up). These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as `NA`)  
* **date**: The date on which the measurement was taken in YYYY-MM-DD format  
* **interval**: Identifier for the 5-minute interval in which measurement was taken  

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. 

## Loading and preprocessing the data

### Downloading and loading the data to RStudio

```{r}
fileURL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if (!file.exists('./repdata data activity.zip')){
        download.file(fileURL, './repdata data activity.zip', mode = 'wb')
        unzip("repdata data activity.zip", exdir = getwd())
}

activitydata <- read.csv("activity.csv", sep = ",", header = T)
```
### Processing/transforming the data into a format suitable for analysis

```{r}
totalsteps <- aggregate(steps ~ date, data = activitydata, sum, na.rm = T)
```

## What is mean total number of steps taken per day?

### Making a histogram of the total number of steps taken each day

```{r}
hist(totalsteps$steps, col = "slateblue4", main = "Histogram of Total Steps taken per day", xlab = "Total steps taken per day", cex.axis = 1, cex.lab = 1)
```

### Calculating and reporting the mean and median total number of steps taken per day

```{r}
meansteps <- mean(totalsteps$steps)
mediansteps <- median(totalsteps$steps)

meansteps
mediansteps
```

* The mean total number of steps taken per day is 10766 steps.  
* The median total number of steps taken per day is 10765 steps.

## What is the average daily activity pattern?

### Making a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)


```{r}
stepsinterval <- aggregate(steps ~ interval, data = activitydata, mean, na.rm = TRUE)

plot(steps ~ interval, data = stepsinterval, type = "l", xlab = "Time Intervals (5-minute)", ylab = "Mean number of steps taken (all days)", main = "Average number of steps taken at 5-minute intervals", col = "slateblue4")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxstepinterval <- stepsinterval[which.max(stepsinterval$steps), "interval"]

maxstepinterval
```

* 835th interval contains the maximum number of steps.

## Imputing the missing values

### Calculating and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missingrows <- sum(!complete.cases(activitydata))

missingrows
```

* The total number of rows with NAs is 2304.

### Devising a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The strategy that is used in this case is to fill all the missing values with the mean step for a given 5-minute interval.

```{r}
meanstepsperinterval <- function(interval){
    stepsinterval[stepsinterval$interval == interval, "steps"]
}
```

### Creating a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
completeactivity <- activitydata

flag = 0
for (i in 1:nrow(completeactivity)) {
        if (is.na(completeactivity[i, "steps"])) {
                completeactivity[i, "steps"] <- meanstepsperinterval(completeactivity[i, "interval"])
                flag = flag + 1
        }
}
```

* Total of 2304 missing values were filled.

### Making a histogram of the total number of steps taken each day.

```{r}
totalstepsperday <- aggregate(steps ~ date, data = completeactivity, sum)

hist(totalstepsperday$steps, col = "slateblue4", xlab = "Total Number of Steps", main = "Histogram of Total Number of Steps Taken Each Day")
```

### Calculating and reporting the mean and median total number of steps taken per day

```{r}
meanstepsfilled <- mean(totalstepsperday$steps)
medianstepsfilled <- median(totalstepsperday$steps)

meanstepsfilled
medianstepsfilled
```

* Mean total number of steps taken per day is 10766.
* Median total number of steps taken per day is 10766.

### Do these values differ from the estimates from the first part of the assignment?

* The *mean* value is the same as the value before imputing missing data, but the *median* value has changed.

### What is the impact of imputing missing data on the estimates of the total daily number of steps?

* The *mean* value is the same as the value before imputing missing data since the mean value has been used for that particular 5-min interval. The median value is different, since the median index is now being changed after imputing missing values.

## Are there differences in activity patterns between weekdays and weekends?

### Creating a new factor variable in the dataset with two levels - “weekday" and "weekend” indicating whether a given date is a weekday or weekend day.

```{r}
completeactivity$day <- ifelse(as.POSIXlt(as.Date(completeactivity$date))$wday%%6 == 0, "Weekend", "Weekday")

completeactivity$day <- factor(completeactivity$day, levels = c("Weekday", "Weekend"))
```

### Making a panel plot containing a time series plot (i.e. type = "l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
stepsintervalfilled = aggregate(steps ~ interval + day, completeactivity, mean)

library(lattice)

xyplot(steps ~ interval | factor(day), data = stepsintervalfilled, aspect = 0.5, type = "l")
```
